Open-source landscape for 3D CSEM modelling

### 2020-04-10 manuscript/Manuscript.tex
Code section as draft

\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
\usepackage{siunitx}                % SI conform commands (eg \num)
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{xspace}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Write at the end; one or two sentence(s) each:
%
\begin{enumerate}
  \item Principal OBJECTIVES AND SCOPE OF THE WORK
  \item Methodology
  \item Results
  \item Conclusions
\end{enumerate}
%
\end{abstract}

@@

The four codes under consideration are, in alphabetical order, \custem
\citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem
\citep{GJI.19.CastilloReyes}, and \simpeg \citep{CAG.15.Cockett}. All four
codes have their user-facing routines written in Python, while the underlying
heavy computation is mostly done in other languages (which is the normal case
in the scientific Python stack); all of them make heavy use of \texttt{NumPy}
\citep{CSE.11.VanDerWalt} and \texttt{SciPy} \citep{NM.20.Virtanen}. All are
“modern” open-source projects, meaning that they not only come with open-source
license, but they are also hosted in an online-hosted version-control system
with feedback and tracking possibilities (raising issues, filing pull
requests), have extensive online documentation including many examples, have
continuous integration to some degree, and also importantly are all easily
installable, without any user-side compilations.

In the following a quick rundown of the codes. It is, however, beyond the scope
of this article to go into every detail of the different modellers, and we
refer to their references and documentation.

\tabl[btp]{codecomparison}{All codes solve Maxwell's Equations in its
  differential form (DE) and with the quasistatic approximation. We use two
  finite element codes and two finite difference codes; the finite element
  codes both make use of the FEniCS solver \citep{CSE.15.Alnaes}, whereas
  \simpeg can use either solvers implemented in SciPy or the PARDISO solver
  \citep{FGCS.04.Schenk}; \emg3d, on the other hand, is a solver on its own.}{
  \centering
  \small
\begin{tabular}{lcccc}
  \toprule
  %
  & \custem & \emg3d & \petgem & \simpeg  \\
  \midrule
  %
  Home & \href{https://custem.rtfd.io}{custem.rtfd.io}
       & \href{https://empymod.github.io}{empymod.github.io}
       & \href{http://petgem.bsc.es}{petgem.bsc.es}
       & \href{https://docs.simpeg.xyz}{simpeg.xyz} \\
  Installation & \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda} \\
  License & GPL-3.0 & Apache-2.0 & GPL-3.0 & MIT \\
  Comp. Dom. & frequency & frequency & frequency & frequency \& time \\
  Method & FE & FIT & FE & FV \\
  Mesh & tetrahedral & rectilinear & tetrahedral & rectilinear; curvilinear \\
  BC & ? & PEC & ? & ? \\
  Solver & FEniCS$^\dagger$ & \emg3d & FEniCS$^\dagger$ &
           PARDISO$^\ddagger$; SciPy \\
  %
  \bottomrule
\end{tabular}}%
%

\imycom[DW]{
Each code should be briefly summarized by four points:\\
- Brief intro and main features\\
- Main selling point\\
- Planned features\\
- List (same list for all)\\
}

\subsection{custEM}

@@

The 3D CSEM modeller \emg3d \citep{JOSS.19.Werthmuller} is a multigrid
\citep{CMMP.64.Fedorenko} solver for 3D electromagnetic diffusion following
\cite{GP.06.Mulder}, with tri-axial electrical anisotropy and isotropic
magnetic permeability. The matrix-free solver can be used as main solver or as
preconditioner for one of the Krylov subspace methods implemented in SciPy
\citep{NM.20.Virtanen}, and the governing equations are discretized on a
staggered grid by the finite-integration technique \cite{AEU.77.Weiland}, which
can be viewed as a finite-volume generalization of \cite{IEEE.66.Yee}'s scheme.
The code is written completely in Python using the NumPy
\citep{CSE.11.VanDerWalt} and SciPy stacks, where the most time- and
memory-consuming parts are sped up through jitted functions using Numba
\citep{LLVM.15.Lam}.

The multigrid method is characterized by almost linear scaling both in terms
of CPU usage and RAM, and it is therefore a comparably low-memory consumption
solver (see results). However, the way multigrid is implemented also means some
additional constraints on the chosen grid. You can input any three-dimensional
grid into \emg3d, but the implemented method works with the existing nodes,
meaning there are no new nodes created as coarsening is done by combining
adjacent cells. The more times the grid dimension can be divided by two the
better it is suited. Ideally, the dimension of the coarsest grid should be a
low prime, and grid dimensions are therefore given by their multiplication with
powers of two (${2,3,5}·2^n$).

Currently \emg3d can be used to model the electric and magnetic fields due to
arbitrary finite length electric dipole sources for a single frequency at a
time. Helper functions to compute time-domain responses are in the works
(should be ready by publishing this article). Current development is focused
on adding inversion capabilities; averaging routines to implement strong
topography; and general anisotropy.

The 3D solver \emg3d can be found in the GitHub organization of \empymod
(`empymod.github.io <https://empymod.github.io>`_). The modeller \empymod
\citep{GEO.17.Werthmuller} is a 1D code which can compute electric or
magnetic responses due to a 3D electric or magnetic source in a layered-earth
model with vertical transverse isotropic (VTI) resistivity, VTI electric
permittivity, and VTI magnetic permeability, from very low frequencies (DC) to
very high frequencies (GPR).

In the future it should be possible to use \emg3d as a solver within the
\simpeg framework.

@@

\clearpage  % TODO: Just for now while drafting, remove later!
\section{Comparison}

@@

>>>>

\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
\usepackage{siunitx}                % SI conform commands (eg \num)
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage{flafter}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
\emph{A will draft a first abstract after the manuscript has been reviewed by
everyone once.}
%
% % Principal objectives and scope of the work
% % Methodology
% % Results
% % Conclusions
%
\end{abstract}

@@

The four codes under consideration are, in alphabetical order, \custem
\citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem
\citep{GJI.19.CastilloReyes}, and \simpeg \citep{CAG.15.Cockett}. All four
codes have their user-facing routines written in Python; all of them make heavy
use of \texttt{NumPy} \citep{CSE.11.VanDerWalt} and \texttt{SciPy}
\citep{NM.20.Virtanen}. All are “modern” open-source projects, meaning that
they not only come with an open-source license, but they are also in an
online-hosted version-control system with feedback and tracking possibilities
(raising issues, filing pull requests), have extensive online documentation
including many examples, have continuous integration to some degree, and also
importantly are all easily installable, without any user-side compilations.

In the following a quick rundown of the codes. It is, however, beyond the scope
of this article to go into every detail of the different modellers, and we
refer to their documentations for more details. An overview comparison of the
different codes is given in Tabel~\ref{tbl:codecomparison}. All codes have in
common that they solve the Maxwell's equation in its differential form (DE)
under the quasistatic or diffusive approximation, hence neglecting displacement
currents.

\tabl[btp]{codecomparison}{All codes solve Maxwell's equations in its
  differential form (DE) under the quasistatic approximation. Note that
  \emg3d is a solver on its own, while the other codes implement third-package
  solvers such as PETSc, MUMPS, FEniCS \citep{CSE.15.Alnaes}, or PARDISO
  \citep{FGCS.04.Schenk}.
  \imycom[DW]{PETSc and MUMPS probably need citations too.}}{
  \centering
  \footnotesize
\begin{tabular}{lcccc}
  \toprule
  %
  & \custem & \emg3d & \petgem & \simpeg  \\
  \midrule
  %
  Home & \href{https://custem.rtfd.io}{custem.rtfd.io}
       & \href{https://empymod.github.io}{empymod.github.io}
       & \href{http://petgem.bsc.es}{petgem.bsc.es}
       & \href{https://docs.simpeg.xyz}{simpeg.xyz} \\
  License & GPL-3.0 & Apache-2.0 & GPL-3.0 & MIT \\
  Installation & \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda} \\
  Comp. Dom. & frequency & frequency & frequency & frequency \& time \\
  Method & FE & FIT & FE & FV \\
  Mesh & tetrahedral & rectilinear & tetrahedral & rectilinear; curvilinear \\
  BC & Dirichlet & PEC & Neumann; Dirichlet & Dirichlet; Neumann \\
  Solver & PETSc; MUMPS & \emg3d & FEniCS & PARDISO; MUMPS \\
  %
  \bottomrule
\end{tabular}}%
%

\imycom[DW]{
Each code should be briefly summarized by three points:\\
- Brief intro\\
- Main selling point\\
- Planned features\\
I think in a first round \textbf{everyone should just write what they want to
write and think is important for their code}. In a second round we should then
homogenize it, such that all codes have more or less the same length and also
similar topic/layout.
}

\clearpage  % TODO: Just for now while drafting, remove later!
\subsection{custEM}

@@

The 3D CSEM modeller \emg3d is a multigrid \citep{CMMP.64.Fedorenko} solver for
3D electromagnetic diffusion following \cite{GP.06.Mulder}, with tri-axial
electrical anisotropy, isotropic electric permittivity, and isotropic magnetic
permeability. The matrix-free solver can be used as main solver or as
preconditioner for one of the Krylov subspace methods implemented in SciPy, and
the governing equations are discretized on a staggered grid by the
finite-integration technique \citep{AEU.77.Weiland}. The code is written
completely in Python using the NumPy and SciPy stacks, where the most time- and
memory-consuming parts are sped up through jitted (just-in-time compiled)
functions using Numba \citep{LLVM.15.Lam}. The code computes the electric field
due to an electric source in the frequency-domain. Implemented are functions to
obtain the magnetic field due to an electric source; to obtain the electric and
magnetic fields due to a magnetic source; and to obtain time-domain responses.

The multigrid method is characterized by almost linear scaling both in terms of
runtime (CPU) and memory (RAM) usage, and it is therefore a comparably
low-memory consumption solver. It is also minimal in terms of requirements,
only NumPy, SciPy, and Numba are required, with Python 3.7+. The current
development is focused on adding inversion capabilities. A further plan is also
to implement a hook in order to use \emg3d as a solver for \simpeg.

@@

\clearpage  % TODO: Just for now while drafting, remove later!
\section{Verifications}

@@

=> \textbf{Important point: Verification of existing and future codes: this is
a start!}
